Trade-off between market and ecosystem services drives settlement decisions among smallholder ranchers in Baja California Sur, Mexico

### 2024-01-25 manuscript/manuscript.qmd
remove multiple bib entries for same reference

To explore this issue, we focus on a small group of historically marginalized
rural ranchers living in the arid Sierra de la Giganta region of Baja California
Sur (BCS), Mexico [@macfarlan2020a, see @fig-overview-map]. We first explore how
market integration and local ecology structure the distribution of occupied and
abandoned ranch clusters. This is a process that took place over several hundred
years, with the shift to its current distribution beginning in the late 1970s,
following construction of the transpeninsular Federal Highway 1 [@barkenbus1974;
@mathes2015]. We then ask how market integration and local ecology have shaped
the distribution of the population across occupied clusters today.

@@

Given the aridity of the area and lack of regionally connective water
infrastructure, rancheros depend almost entirely on surficial springs to satisfy
their water needs, including drinking water for themselves and their livestock,
as well as irrigation for garden plots (known as huertas and jardines), which
provide a small supplement to human and livestock diets during dry periods
[@macfarlan2020a; @macfarlan_etal2020a; @schniter2021; @lerback2022]. To make
use of springs, ranchers either have to pump water themselves, using gas or
solar powered equipment, or transport it by gravity irrigation [@lerback2022]. A
strong incentive thus exists to locate in closer proximity to these water
resources.

@@

In BCS (along with the rest of Mexico), the Agrarian Land Reforms of 1992 played
an outsized role in driving these changes, as they introduced powerful new
incentives for *ejidos* to parcelize collective land holds and allocate private
rights to individuals [@barnes2009]. While note the sole cause, this new
property regime has increased uncertainty and insecurity about land title and
exposed the region to speculation from real estate markets. As has happened
elsewhere, these privatization efforts appear to be led by NGOs, commercial
interests, and other external forces, with the result being greater
fragmentation of pastoral rangeland systems and increased uncertainty about land
use [@burnsilver_etal2008; @galvin2009]. For the rural ranchers of the Sierra de
la Giganta, this process of fragmentation and increasing insecurity at least
partially explains why ranches of related individuals tend to cluster.
Rancheros, unable to find new land for themselves, simply build ranches next to
their kin on land that is at least historically associated with their family.

@@

As mentioned above, this is also an extremely arid region. It is a southern
extension of the Sonoran Desert Biogeographic Region, so the area is dominated
by columnar cacti and xeric scrub [@shreve_wiggins1964;
@leondelaluz_dominguezcadena2006] capable of sustaining themselves in a place
that averages only 200 mm of rainfall every year [@rebman_roberts2012]. This
circumstance is not helped by the lack of surficial freshwater stocks
[@maya1997], with the only permanent sources of naturally occurring freshwater
being arid-land springs [@grismer1993; @macfarlan2020a; @macfarlan_etal2020a;
@macfarlan2021]. @lerback2022 provides a fine-grained analysis of these spring
systems in the project area, describing them as primarily lentic pools at low
points in the arroyos (see @fig-context-photos B).

@@

Our first statistical model indicates that abandoned ranches were located
farther from springs. Although this appears to run counter to our previous claim
about the importance of springs to all settlement in the region, we believe this
is an artifact of our analytic framework, which required us to collapse the
temporal dimension of this largely historical process. Historical and
ethnographic evidence suggest that as the first Euro-American colonists entered
this region in the 18th century, they occupied the largest springs in close
proximity to arable land [@crosby1994; @macfarlan_etal2020;
@macfarlan_etal2021]. As the population expanded over time, individuals would
move up or down drainages, to the next closest spring that could support
subsistence needs. After time, ranchers likely saturated the landscape by
occupying all areas with permanent freshwater [sensu @crosby1981] as well as
less desirable locations which were supported not by springs, but by ephemeral
freshwater sources such as tinajas and seasonally filled lake basins, which can
hold water for months or years if they receive sufficient precipitation via the
North American Monsoon.

>>>>

To explore this issue, we focus on a small group of historically marginalized
rural ranchers living in the arid Sierra de la Giganta region of Baja California
Sur (BCS), Mexico [@macfarlan_etal2020b, see @fig-overview-map]. We first
explore how market integration and local ecology structure the distribution of
occupied and abandoned ranch clusters. This is a process that took place over
several hundred years, with the shift to its current distribution beginning in
the late 1970s, following construction of the transpeninsular Federal Highway 1
[@barkenbus1974; @mathes2015]. We then ask how market integration and local
ecology have shaped the distribution of the population across occupied clusters
today.

@@

Given the aridity of the area and lack of regionally connective water
infrastructure, rancheros depend almost entirely on surficial springs to satisfy
their water needs, including drinking water for themselves and their livestock,
as well as irrigation for garden plots (known as huertas and jardines), which
provide a small supplement to human and livestock diets during dry periods
[@macfarlan_etal2020b; @macfarlan_etal2020a; @schniter2021; @lerback2022]. To
make use of springs, ranchers either have to pump water themselves, using gas or
solar powered equipment, or transport it by gravity irrigation [@lerback2022]. A
strong incentive thus exists to locate in closer proximity to these water
resources.

@@

In BCS (along with the rest of Mexico), the Agrarian Land Reforms of 1992 played
an outsized role in driving these changes, as they introduced powerful new
incentives for *ejidos* to parcelize collective land holds and allocate private
rights to individuals [@barnes2009]. While not the sole cause, this new property
regime has increased uncertainty and insecurity about land title and exposed the
region to speculation from real estate markets. As has happened elsewhere, these
privatization efforts appear to be led by NGOs, commercial interests, and other
external forces, with the result being greater fragmentation of pastoral
rangeland systems and increased uncertainty about land use
[@burnsilver_etal2008; @galvin2009]. For the rural ranchers of the Sierra de la
Giganta, this process of fragmentation and increasing insecurity at least
partially explains why ranches of related individuals tend to cluster.
Rancheros, unable to find new land for themselves, simply build ranches next to
their kin on land that is at least historically associated with their family.

@@

As mentioned above, this is also an extremely arid region. It is a southern
extension of the Sonoran Desert Biogeographic Region, so the area is dominated
by columnar cacti and xeric scrub [@shreve_wiggins1964;
@leondelaluz_dominguezcadena2006] capable of sustaining themselves in a place
that averages only 200 mm of rainfall every year [@rebman_roberts2012]. This
circumstance is not helped by the lack of surficial freshwater stocks
[@maya1997], with the only permanent sources of naturally occurring freshwater
being arid-land springs [@grismer1993; @macfarlan_etal2020b;
@macfarlan_etal2020a; @macfarlan_etal2021]. @lerback2022 provides a fine-grained
analysis of these spring systems in the project area, describing them as
primarily lentic pools at low points in the arroyos (see @fig-context-photos B).

@@

Our first statistical model indicates that abandoned ranches were located
farther from springs. Although this appears to run counter to our previous claim
about the importance of springs to all settlement in the region, we believe this
is an artifact of our analytic framework, which required us to collapse the
temporal dimension of this largely historical process. Historical and
ethnographic evidence suggest that as the first Euro-American colonists entered
this region in the 18th century, they occupied the largest springs in close
proximity to arable land [@crosby1994; @macfarlan_etal2020b;
@macfarlan_etal2021]. As the population expanded over time, individuals would
move up or down drainages, to the next closest spring that could support
subsistence needs. After time, ranchers likely saturated the landscape by
occupying all areas with permanent freshwater [sensu @crosby1981] as well as
less desirable locations which were supported not by springs, but by ephemeral
freshwater sources such as tinajas and seasonally filled lake basins, which can
hold water for months or years if they receive sufficient precipitation via the
North American Monsoon.
